---
title: "E6690: Statistical Learning for Bio & Info System Final Project - Stop-and-Frisk in New York City"
author: "Lillian Chik, Mun Shin (CVN - mc3549)"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Desktop/E6690_mc3549_FinalProject/Data")
library(kernlab)
data(spam)
```

# 1. Introduction

Stop-and-frisk program is a policy in New York City where police officers can stop, detain and search a person if they suspect he carries a weapon or other forms of contraband. Every time a police officer stopped a suspect, he must document the stop by completing a UF-250 stop-and-frisk form (Goel et al. 368). The data we used derived from this form, and it includes information about the suspect’s ethnicity, location, time of stop, reason of stop and whether the suspect is eventually being searched. On this paper, we performed statistical analysis such as t-test, linear regression, logistic regression, Bayesian model, and quantile-quantile plotting. From the result, we provide reasoning of why the stop-and-frisk program is not effective in reducing crimes among civilians in New York City.  

# 2. Original Data Set and Paper

On this analysis, I used Goel et al.’s paper as my primary source. Using linear regression, they analyzed the stop-and-frisk data from January 1st, 2008 to December 31th, 2012 to determine whether racial bias placed a role in influencing police officers’ decision to stop, search and frisk a pedestrian (Goel et al. 368). According to Goel et al., Blacks and Hispanics are more likely to be searched than Whites when being compared to their actual population proportion, even though they are less likely to be found with a weapon (379). They found evidence that the criteria for police officers to search Blacks and Hispanics are much less stringent compared to Whites (Goel et al. 387). Finally, they found that if the police officers have conducted only 6% of the stops that has the highest probability of finding a weapon, then they will recover most of the contrabands on the streets and thus decrease the number of unnecessary stops among all the pedestrians in New York City (Goel et al. 382). At the end, Goel et. al. also gives recommendations on how to improve the methods of collecting the data that was self-reported by police officers (388). We adapted Goel et. al.’s approaches and examined a few questions to determine the effectiveness of the stop-and-frisk policy and whether or not it is a form of racial profiling against certain minority groups in New York City. For this analysis, we used the stop-and-frisk data released by the New York City police department. It contains a total of 3.8 million documented stops from 2003 to 2017. The data is in the form of large csv files in Excel, and each column has a piece of information describing the details of a particular stop. All the columns in the data followed the same format as the UF-250 stop-and-frisk form, and approximately 120 features are recorded. 

# 3. Reproduce
```{r, echo = FALSE, eval = TRUE}
sqf2010 = read.csv("2010.csv")
sqf2015 = read.csv("2015.csv")
```
In the stop-question-frisk (SQF) policy, because the police officers are responsible for self-report instances of suspected crimes, it raised the question of whether or not the data police officers have recorded is accurate. In particular, we are interested to understand if the police officers have identified the suspect’s ethnicity correctly, because mistaken identity could have a huge impact in affecting the accuracy of the data. In this analysis, we only looked at a selective sample of the data where the suspects are documented by police officers as Asians or Latinos. Using conditional probability, we assumed the probability that a police officer has correctly identify an Asian is 0.95, otherwise the Asian suspect is identified incorrectly as Latinos. We also assumed that the probability that a police officer has correctly identify a Latino suspect is 0.95, otherwise the Latino suspect is identified incorrectly as Asian. Using these assumptions, we then calculate the probability that a suspect who is subject to stops and searches is Asian, and the probability that a suspect who is subject to stops and searches is Latino using the 2010 SQF data. This problem has two equations (one for Asian and one for Latino) and two unknown variables (the actual percentage of Latino and Asian suspect in the New York City population). Using a set of formulas, we calculated the probability that a suspect who was being stopped is Latino is 0.906 and the probability that a suspect who was being stopped is Asian is 0.094. From there, we used a set of conditional probability formulas with Bayes’ Theorem and calculated the probability that a police officer has identified a suspect correctly as Asian is 0.494 with false positive of 0.506, and the probability that a police officer has identified a suspect correctly as Latino is 0.997 with false positive of 0.03. Going back to the initial assumption, it means 0.05 of the Latino suspects the police officers have identified incorrectly is consists of 0.506 of the Asian stopped suspects in the data, and 0.05 of the Asian suspects the police officers have identified incorrectly is consists of 0.03 of the Latino stopped suspects in the 2010 SQF data. The results indicated that the data by ethnicity is not very reliable especially for the subset of Asian stopped suspects. 

**Figure 1. Table describes all the stops made by police officers for each ethnicity in 2010**  
```{r, echo = FALSE, eval = TRUE}
table(sqf2010$race)
```

Using the table, we know there is 19732 Asian/Pacific Islander suspects, 38689 Black-Hispanic suspects, and 150637 White-Hispanic suspect. From there, the total number of Hispanic and Asian/Pacific Islander suspect is 209058 using summation. 

Using Bayes' Theorem in conditional probability, we can calculate the following: 

Let  
Event $I$: a suspect is identified as Hispanic = 0.906  
Event $H$: a suspect is actually Hispanic = 0.094  
Event $I^c$: a suspect is identified as not Hispanic (i.e. suspect is identified as Asian/Pacific Islander) = 0.997  
Event $H^c$: a suspect is not actually Hispanic (i.e. suspect is actually Asian/Pacific Islander) = 0.494  

Then  
The probability that a stopped suspect is Hispanic: $P(I)$  
The probability that a stopped suspect is Asian/Pacific Islander: $P(I^c)$  
The probability that someone who is identified by the officer as Hispanic is actually Hispanic: $P(H|I)$  
The probability that someone who is identified as Asian/Pacific Islander is actually Asian/Pacific Islander: $P(H^c|I^c)$  


$P(I) = \frac{38689+150637}{38689+150637+19732} = 0.906$  
$P(I^c) = \frac{19732}{38689+150637+19732} = 0.094$  

We know that a Hispanic person has a 95\% chance of correctly being identified as Hispanic, and otherwise is misidentified as Asian/Pacific Islander, so  
$P(I|H) = 0.95$  

We know that an Asian/Pacific Islander has a 95\% chance of correctly being identified as Asian/Pacific Islander, and otherwise is misidentified as Hispanic, so  
$P(I^c|H^c) = 0.95$  

From the Law of Total Probability, we get  
$P(H) = P(H|I)P(I) + P(H|I^c)P(I^c)$  

Using Bayes’ Rule, we solve for  
$P(H|I) = \frac{P(I|H)P(H)}{P(I)}$  

Plug in P(H):  
$P(H|I) = \frac{P(I|H) [P(H|I)P(I) + P(H|I^c)P(I^c)]}{P(I)}$  
Plug in values:	 
$P(H|I) = \frac{0.95 * [ P(H|I) * 0.906 + P(H|I^c) * (0.094)]}{0.906}$		(1)  

Using Bayes’ Rule, we solve for  
$P(H^c|I^c) = \frac{P(I^c|H^c)P(H^c)}{P(I^c)}$  
$1-P(H|I^c) = \frac{P(I^c|H^c) [1-P(H)]}{P(I^c)}$  

Plug in P(H) and values:  
$1-P(H|I^c) = \frac{0.95 * [1-(P(H|I) * 0.906 + P(H|I^c) * 0.094)]}{0.094}$    	(2)  

Solving (1) and (2):  
$P(H|I) = \frac{4066}{4077} = 0.997$  
$P(H|I^c) = \frac{214}{423} = 0.506$  
$P(H^c|I^c) = 1 - P(H|I^c) = 1 - 0.506 = 0.494$  

On the second analysis, we wanted to follow Goel et. al.’s approaches to examine the demographic factors such as ethnicity and borough and see if they play a significant role on the percentage an individual who was stopped by police officers in the 2010 SQF data. For the first part, we used linear regression model to compare the percentage that a person was being searched in all the ethnicities using the subset of Asian stopped suspect as the reference measure. By comparing the percentage an individual who was stopped is identified as Asian and the percentages a stopped suspect is in the other ethnicities, it showed that most subsets are indeed have a high significant of being stopped because their respective p-values are less than 0.05. But Unknowns and American Indian/Alaskan Natives subsets are low significant because their respective p-values are more than 0.05 when comparing its respective percentage difference to the percentage an individual who was stopped is identified by police officers as Asian. Whites have the lowest estimates of -0.022294 and Black-Latinos has the highest estimate of 0.161044, it means Whites have the lowest probability of being searched while Black-Latinos have the highest probability of being searched. 

**Figure 2. Table describes the number of suspects being searched by each ethnicities**  
```{r, echo=FALSE, eval=TRUE}
table(sqf2010$frisked,sqf2010$race) #table that compared searched rates for each ethnicities
```

**Figure 3. Linear regression model to compare the searched rates of all ethnicities using Asian/Pacific islander used as baseline:**  
```{r, echo=FALSE, eval=TRUE}
sqf2010$ethnicity <-relevel(sqf2010$race, "ASIAN/PACIFIC ISLANDER")
regressionModelRace<- lm(frisked ~ ethnicity, data = sqf2010) 
summary(regressionModelRace)
```

On the second part, we used linear regression to compare the percentage that a person was being searched among all the boroughs compared to the percentage in Manhattan as reference measure. Looking at the p-values, the result showed that most subsets are indeed high significant because their respective p-values are less than 0.05. Meanwhile, Brooklyn and Manhattan have the lowest percentages when comparing the percentages of being stopped among all boroughs. Ranking the estimates from lowest to highest (Staten Island, Manhattan, Brooklyn, Queens, and Bronx), these results support the claim that Bronx has the highest crime rate in all of the borough with Black-Latinos have the highest chance subject to being stopped and searched in New York City. These results are consistent with Goel et. al.’s findings. 

**Figure 4. Table describes the searched rate by each borough:**  
```{r, echo=FALSE, eval=TRUE}
table(sqf2010$frisked,sqf2010$city) #table that compared searched rates for each borough
```

**Figure 5. Linear regression model to compare searched rates of all boroughs using Manhattan as baseline:**  
```{r, echo=FALSE, eval=TRUE}
sqf2010$borough <-relevel(sqf2010$city, "MANHATTAN") #
regressionModelCity<- lm(frisked ~ borough, data = sqf2010) 
summary(regressionModelCity)
```

We also want to compare if there is significance difference between the percentages an individual is being searched, because of his gender and age. To do this, we used 2010 SQF data and plotted the age histogram and Q-Q plot for both male and female stopped suspects. We want to determine if the male and female have equal likelihood of being stopped among all ages. Looking at the result, both histograms have a normal distribution and similar in shape. But they both have positive skewness with the mean age greater than the medium age. It means the female and male suspects who are being stopped are mostly young adults. On the Q-Q plots, the line is skewed by its edges because some of the data in the age column are 0 or 999 and police officers are unsure about the age of the suspect they reported. 

**Figure 6-9. Histograms and QQ Plots describes the distribution of ages among male and female suspects**  
```{r, echo=FALSE, eval=TRUE}

# Separate the data for ages into male and female sections
agesForMales <- sqf2010$age[sqf2010$sex == "M"]
agesForFemales <- sqf2010$age[sqf2010$sex == "F"]

# Clean out obviously wrong values
agesForMales <- agesForMales[agesForMales > 0 & agesForMales < 100]
agesForFemales <- agesForFemales[agesForFemales > 0 & agesForFemales < 100]

par(mfrow=c(2,2))
hist(agesForMales, main = "Ages Of Male Suspects")
hist(agesForFemales, main = "Ages Of Female Suspects")

# Create Q-Q Normal Plots of the two sets of data to determine whether or not
# they are normally distributed
qqnorm(agesForMales, main = "Normal Q-Q Plot for Male Suspect Ages")
qqline(agesForMales, distribution = qnorm)
qqnorm(agesForFemales, main = "Normal Q-Q Plot for Female Suspect Ages")
qqline(agesForFemales, distribution = qnorm)
```

We also used Q-Q plots to compare the age distribution of the male suspects to all the population and the age distribution of the female suspect to the male suspects. Although the number of male suspect being searched are much higher in ratio compared to the number of female suspects, but their age distributions are still very similar to each other and their Q-Q plots can be considered as linear. From this, it means that gender is not an influencing factor that contribute to the likelihood a person will be frisked by police officers. 

**Figure 10-11. QQ plots comparing the ages of male suspects to the ages of female suspects, and comparing the ages of male suspects to the ages of total population**  
```{r, echo=FALSE, eval=TRUE, out.width='300px'}
par(mfrow=c(1,2), ps=10)
# Create a Q-Q plot to compare them
qqplot(agesForFemales, agesForMales, main = "Q-Q Plot of Male/Female Ages")

# Compare ages of male suspects to that of the entire population
allAges <- sqf2010$age[sqf2010$age > 0 & sqf2010$age < 100]
qqplot(allAges, agesForMales, main = "Q-Q Plot of Male/All Ages")
```

## 4. New Approach

For this new analysis, we want to find out the probability that a person is subject to stops and searches for the total population and for those who did not want to identify themselves at a 0.95 confidence interval in 2015. We want to compare the difference between these two probabilities to understand if the likelihood of being arrested will increase if the suspect is refused to identify himself while the police officers are detaining him. 

**Figure 12. Table comparing searched rate for the different type of identification suspect provided to the police officer:**  
``` {r, echo = FALSE, eval = TRUE}
table(sqf2015$frisked, sqf2015$typeofid)
```

**Entire population:**  

$\hat{p} = \frac{15257}{7306 + 15257} =$ `r  15257/(7306 + 15257)`
$n = 22563$  

*95\% confidence interval:*  
$(\hat{p} - 1.96 * \sqrt{\frac{\hat{p}(1-\hat{p})}{n-1}}, \hat{p} + 1.96 * \sqrt{\frac{\hat{p}(1-\hat{p})}{n-1}})$  
$(0.6701, 0.6823)$  
 
**Refused to give identification:**  

$\hat{p} = \frac{390}{249 + 390} =$ `r  390/(249+390)`
$n = 639$  

*95\% confidence interval:*  
$(\hat{p} - 1.96 * \sqrt{\frac{\hat{p}(1-\hat{p})}{n-1}}, \hat{p} + 1.96 * \sqrt{\frac{\hat{p}(1-\hat{p})}{n-1}})$  
$(0.5725, 0.6482)$  

Looking at the results above, the probability that a person is subject to stops and searches is between 0.6701 and 0.6823 for the total population at a 0.95 confidence interval. But when the sample used for the analysis changes from total population to a subset of individuals who refuse to identify themselves, the probability that a person is subject to stops and searches decreases to between 0.5725 and 0.6482. This seems to suggest that a person who refused to provide an ID is less likely to be stopped by the police officers than the entire population. To further support this claim, we performed t-test to compare the difference between the probabilities of these two cases. At a 0.95 confidence interval, the difference estimates indicate that the probability that a person is subject to stops and searches for the total population is in the range of 0.07680032 greater than the probability that a person is being searched after refusing to identify himself. This result is consistent with my previous analysis in this problem, and infers that people who refused to identify themselves have a lesser probability to be searched than the total population. 

**Figure 13. T-test to examine the difference of searched rates between the percentage of suspects frisked for the entire population and the percentage of suspects frisked for the subset who refused to give their identification**  
```{r, echo=FALSE, eval=TRUE}
friskedTotalPopulation <- sqf2015$frisked
friskedSubsetSuspectsRefusedID <- sqf2015$frisked[sqf2015$typeofid == "REFUSED"]
t.test(x=friskedTotalPopulation, y=friskedSubsetSuspectsRefusedID, conf.level=0.95)
```

For the other new analysis, we want to understand which particular binomial features have a big impact on the length of stopped time made by police officers. At first, we used linear regression and treated the length of stopped time as the dependent variable and the binomial features as the independent variables. We also revised the data such that it removed the outlier points on the length of stopped time parameter. Looking at the figures, the result seems to suggest that the residuals of the regression model has a normal distribution. Using this, we can confirmed that the p-values are accurate in the model. Looking at the p-values, all the binomial features have p-values less than 2E10-16 except the presence of weapon. It means we can not reject the null hypothesis that the presence of weapon and length of stopped time are independent from each other.    

**Figure 14. Logistic regression for length of stops against binary variables: `arstmade`, `searched`, `inside`, `sumissue`, `frisked`, `weap`, `contrabn`, `radio`, and `pf`**  
```{r, echo = FALSE, eval = TRUE}
perstopCleaned <- sqf2010$perstop[sqf2010$perstop < 40 & sqf2010$perstop > 0]
perstopLog <- log(perstopCleaned + 1)
arstmadeCleaned <- sqf2010$arstmade[sqf2010$perstop < 40 & sqf2010$perstop > 0]
searchedCleaned <- sqf2010$searched[sqf2010$perstop < 40 & sqf2010$perstop > 0]
insideCleaned <- sqf2010$inside[sqf2010$perstop < 40 & sqf2010$perstop > 0]
sumissueCleaned <- sqf2010$sumissue[sqf2010$perstop < 40 & sqf2010$perstop > 0]
friskedCleaned <- sqf2010$frisked[sqf2010$perstop < 40 & sqf2010$perstop > 0]
weapCleaned <- sqf2010$weap[sqf2010$perstop < 40 & sqf2010$perstop > 0]
contrabnCleaned <- sqf2010$contrabn[sqf2010$perstop < 40 & sqf2010$perstop > 0]
radioCleaned <- sqf2010$radio[sqf2010$perstop < 40 & sqf2010$perstop > 0]
pfCleaned <- sqf2010$pf[sqf2010$perstop < 40 & sqf2010$perstop > 0]

model <- lm(perstopLog ~ arstmadeCleaned + searchedCleaned + insideCleaned + sumissueCleaned + friskedCleaned + weapCleaned + contrabnCleaned + radioCleaned + pfCleaned, data = sqf2010)
summary(model)
```

**Figure 15-17. Plot, histogram, and Normal QQ-plot describe the residuals on the regression line**  
```{r, echo = FALSE, eval = TRUE}
par(mfrow=c(1,1))
plot(model$residuals, main = "Residuals Plot")
```

```{r, echo = FALSE, eval = TRUE}
par(mfrow=c(1,2))
hist(model$residuals, main = "Residuals Histogram")
qqnorm(model$residuals, main = "Residuals Normal Q-Q Plot")
qqline(model$residuals, distribution = qnorm)
```

Therefore, we performed a second linear regression model without including the presence of weapon feature. Looking at the result, it can be inferred that all the binomial features are high significant on the length of stopped time made by police officers, except the presence of weapon. The residuals on the linear regression model has a normal distribution. The R-squared value is very small, so it means that the binomial features contribute to only 0.05 of the variance in the length of stopped time. The result is logical because there could be many other factors that contribute to the length of stopped time made by police officers besides the binomial features described in the analysis.  

**Figure 18. Logistic regression model for length of stops against binary variables except weap: `arstmade`, `searched`, `inside`, `sumissue`, `frisked`, `contrabn`, `radio`, and `pf`**  
```{r, echo = FALSE, eval = TRUE}
model2 <- lm(perstopLog ~ arstmadeCleaned + searchedCleaned + insideCleaned + sumissueCleaned + friskedCleaned + contrabnCleaned + radioCleaned + pfCleaned, data = sqf2010)
summary(model2)
```

**Figure 19-21. Plot, histogram, and Normal QQ-plot describes the residuals on regression line**  
```{r, echo = FALSE, eval = TRUE}
par(mfrow=c(1,1))
plot(model2$residuals, main = "Residuals Plot")
```

```{r, echo = FALSE, eval = TRUE}
par(mfrow=c(1,2))
hist(model2$residuals, main = "Residuals Histogram")
qqnorm(model2$residuals, main = "Residuals Normal Q-Q Plot")
qqline(model2$residuals, distribution = qnorm)
```

# 5. Discussion and Conclusion

Based on the analysis on 2010 and 2015 SQF data, we support Goel et. al.’s conclusion that the stop-and-frisk program in New York City is not very effective in reducing the crimes among civilians in the city (387). There is 0.52 probability that a suspect who is temporarily stopped by police officers is being searched for weapons and other contraband, but only 0.02 probability that a weapon is found. Black and Latino suspects are also more likely subject to being stops and searches by police officers compared to White suspects, even though they are less likely to be found with a weapon. On the other hand, the stop-and-frisk program do not seem to have a bias toward age and gender, because the analysis showed that the age distribution of both female and male suspect is normal. From these result, we concluded that the stop-and-frisk program is not very effective in reducing crimes in New York City but rather it creates the dialogue of racial profiling in America because Blacks and Latinos are disproportionally stopped and searched by police officers. On the analysis, we used both histograms and Q-Q plots to examine if the data is normally distributed. Although histogram is very simple and useful to visualize the distribution of samples in the data, we found Q-Q plot to be a much better technique because we can visualize how far off the dataset is away from the normal distribution. For future work, the choice of other statistical analysis techniques should be considered besides using linear regression to examine the relationships described previously. Because the dependent variable and independent variables are binomial, it makes sense to use linear regression to understand the relationship between the parameters. But there are some assumptions we made on the use of linear regression. For example, we currently assumed all the binomial features are independent. In this model, we did not examine rigorously on whether or not there is a correlation between each of the binomial feature against each other. We decided to use this method because all the parameters in the data are not continuous, therefore it is difficult to determine its correlation and how far the variance in the sample might be changing due to potential issue of dependence. But because of the population density in New York City, it is very possible that some of the parameters in the data are dependent on each other. If so, this could influence the standard errors and cause the estimates of the regression model to be inaccurate. There is also a non-randomness factor contribute to how this data is being collected at the first place. We assumed the stop-and-frisk data is randomness by nature, where police officers self-reported incidents of suspected crimes. However, it is possible that there could have personal bias in who is interacting with the police and who is being targeted. The data examines whether police officers have a bias after they made the decision to stop, search and detain someone on the street, but it does not address the bias of whether the documented stop should happen at the first place. Therefore, the data suggests there is racial bias in the stop-and-frisk program, and also inferred a non-randomness in how this data is created at the first place, especially because the interactions between police officers and civilians are not very well documented. This issue is very complex and might not be easily solve by statistical techniques. But rather it highlights the question of what is the best approach to collect these data for future studies, and how to correctly record police interactions with the civilians. Therefore, the complexity of this issue is beyond what is presented in this analysis. 

# 6. References

[1] Goel, Sharad and Rao, Justin M. and Shroff, Ravi. _Precinct or Prejudice? Understanding Racial Disparities in New York City's Stop-and-Frisk Policy (March 2, 2015)._ Annals of Applied Statistics, Vol. 10, No. 1, 365–394, 2016. Available at SSRN: https://ssrn.com/abstract=2572718 or http://dx.doi.org/10.2139/ssrn.2572718

[2] *Stop, Question and Frisk Data.* 2003. New York City Police Department. http://www1.nyc.gov/site/nypd/stats/reports-analysis/stopfrisk.page

[3] Bloomberg, Michael, and Amanda Burden. _NYC2010: Results from the 2010 Census : Population Growth and Race/hispanic Composition._ 2011. New York: NYCPlanning, Department of City Planning. https://www1.nyc.gov/assets/planning/download/pdf/data-maps/nyc-population/census2010/pgrhc.pdf

[4] *Stop and Frisk During the Bloomberg Administration.* 2014. New York Civil Liberties Union. http://www.nyclu.org/files/publications/stopandfrisk_briefer_2002-2013_final.pdf 
